
---
  title: "Reproducible Research: Peer Assessment 1"
  author: Sherwin M
  date: September 1, 2020
  output: md_document
---

## Steps for performing the assignment 

1.Code for reading in the dataset and/or processing the data
2.Histogram of the total number of steps taken each day
3.Mean and median number of steps taken each day
4.Time series plot of the average number of steps taken
5.The 5-minute interval that, on average, contains the maximum number of steps
6.Code to describe and show a strategy for imputing missing data
7.Histogram of the total number of steps taken each day after missing values are imputed
8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9.All of the R code needed to reproduce the results (numbers, plots, etc.) in the report


##Step 1: Code for reading in the dataset and/or processing the data
```{r}
  activity<-read.csv("activity.csv")
  str(activity)
  activity$date<-as.Date(activity$date,"%Y-%m-%d")
```


##Step 2: Histogram of the total number of steps taken each day
```{r}
  step_by_day<-aggregate(steps~date,activity,sum)
  hist(step_by_day$steps, main = "Steps taken per Day", xlab = "Steps", col="red")
  dev.copy(png,"plot1.png",480,480)
  dev.off()
```


##Step 3: Mean and median number of steps taken each day
```{r}
  mean_steps_day<-mean(step_by_day$steps)
  median_steps_day<-median(step_by_day$steps)
```

# What is mean total number of steps taken per day?
The mean of the total number of steps taken per day is *10766.19* and the median of the total number of steps taken per day is *10765*.

##Step 4: Time series plot of the average number of steps taken
```{r}
  step_by_interval<-aggregate(steps~interval,activity,mean)
  plot(step_by_interval$interval, step_by_interval$steps, type="l", main = "Average steps over number of days", xlab = "Interval", ylab="Steps")
  dev.copy(png,"plot2.png",480,480)
  dev.off()
```


##Step 5: The 5-minute interval that, on average, contains the maximum number of steps
```{r}
  max_steps_row <- which.max(step_by_interval$steps)
  step_by_interval[max_steps_row,]
```

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
The interval *835* contains the highest number of steps i.e. *206.167*.

##Step 6: Code to describe and show a strategy for imputing missing data
There are multiple strategies to deal with multiple value imputations.
The common strategies include:
1. Constant value imputations
2. Regression model value imputations
3. Mean/mode value substitutions
For the purpose of simplicity, in this question, I will use the mean/mode value substitution strategy to impute missing values. That is, using the mean values to substitute out the missing values in the original data set
Before doing any sort of imputation, it is helpful to understand what are the distributions of missing values by date and interval
```{r}
activity_imputed<-activity
for(i in 1:nrow(activity_imputed)){
  if(is.na(activity_imputed$steps[i])){
    interv<-activity_imputed$interval[i]
    activity_imputed$steps[i]<-step_by_interval$steps[step_by_interval$interval==interv]
  }
}
```


## Step 7: Histogram of the total number of steps taken each day after missing values are imputed
```{r}
  step_by_day<-aggregate(steps~date,activity_imputed,sum)
  hist(step_by_day$steps, main = "Steps taken per Day", xlab = "Steps", col="red")
  dev.copy(png,"plot3.png",480,480)
  dev.off()
```

# Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
  mean_steps_day<-mean(step_by_day$steps)
  median_steps_day<-median(step_by_day$steps)
```
Yes, these values do differ slightly from the first part of the assignment. There is a slight impact on imputing the missing data.


## Step 8: Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
  activity_imputed["day"]=NA
  weekday<-c("Monday","Tuesday","Wednesday","Thursday","Friday")
  weekend<-c("Saturday","Sunday")
  for (i in 1:nrow(activity_imputed)){
    if(weekdays(activity_imputed$date[i]) %in% weekend){
      activity_imputed$day<-"weekend"
    }
    else{
      activity_imputed$day[i]<-"weekday"
    }
  }
  step_by_day<-aggregate(steps~interval+day,activity_imputed,mean)
  library(lattice)
  xyplot(step_by_day$steps ~ step_by_day$interval|step_by_day$day, main="Average 
         Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
  dev.copy(png,"plot4.png",480,480)
  dev.off()
```

# Are there differences in activity patterns between weekdays and weekends?
Seeing from the graphs we are able to notice slight changes in the average number of steps over weekdays and weekends.
